Supplementary Information

### 2021-04-21 doc/manuscript/ForestGEO_treerings.Rmd
finished SER comments in methods & results

Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and seasonal windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate drivers across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each. The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
#LDW: precip c June (1 mo), PET current June (1mo) W=T
#HF: PDF pMay (1mo), PET cMar (1mo) W=T
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
```
Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
Optimal seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.

The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
Optimal seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).

@@

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the primary climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged (**REFS (NEIL)**). 
Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
~~This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.~~ *(2nd sentence moved to methods)*

>>>>

Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and seasonal windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each. The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the most influential window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
#LDW: precip c June (1 mo), PET current June (1mo) W=T
#HF: PDF pMay (1mo), PET cMar (1mo) W=T
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
```
Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites for which both variables were available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
The most influential seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 8).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season months (< 100 mm rainfall / month).
Outside of the tropics, the most influential windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.

Within the temperature group (Fig. `r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
The most influential seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).

@@

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with the most influential seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

**(should I remove this paragraph?)**
While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
First, in determining the primary climate drivers (step 1; Fig. `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged (**REFS (NEIL)**). 
Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs. `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`). 
~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
~~This decision was motivated by the expectation that differences in the most influential climate windows across species in one site would be minimal compared to cross-site differences (cf. Figs. `r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.~~ *(2nd sentence moved to methods)*

### 2021-04-21 doc/manuscript/ForestGEO_treerings_SI.Rmd
finished SER comments in methods & results

traditional_comparison_generic_legend = "Shown are repsonses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis. Left panels show a month-by-month comparison of *beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable from species-level residual chronologies (traditional approach) and from individual-level analysis in *climwin* (approach presented here). Center panels compare the monthly *beta* coefficient estimates, with the dotted line indicating 1:1 correspondence. Finally, the right panels compare *beta* coefficients for the optimal window selected by *climwin*. Error bars indicate standard error of slope estimates. Note that 1:1 correspondence is not necessarily expected. See Appendix 5 for analysis methods and discussion of expected correspondence."

climwin_generic = "Rows correspond to the four different models run: (1) all RW data, (2) RW for the subset of cores for which $DBH could be reconstructed, (3) BAI, and (4)  AGB growth. The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1). Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal seasonal window (second column). The third and fourth columns give values of linear and quadratic terms for each seasonal window. The final column shows the correlation of individual-level residuals to the selected climate variable, with the function fit in *climwin* (not the final fit)."

@@

The Niobrara site sits at the prairie-forest ecotone of the North American Great Plains. 
There, forests occur in cool, moist canyons formed by spring-fed tributaries along the Niobrara River. 
Cores were taken from seven north-facing *Betula papyrifera* stands along a 27 km section of the Niobrara River, as detailed in @bumann_assessing_2019. 
Niobrara is near the southern limit of *Betula papyrifera*'s distribution, and the population at this site are small and highly disjunct.
The largest healthy trees within each stand were selected for coring.

@@

Prior to analysis, data were prepared and cleaned as described in the Methods section, resulting in an identical set of records for input into each analysis.
For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and seasonal windows can be compared without climwin [e.g., @meko_seascorr_2011], and *climwin* can also be used on residual chronologies. 
Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.
*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the seasonal window identified as optimal by *climwin*.

@@

Finally, we generated plots comparing month-by-month *beta* coefficients describing climate sensitivity, and also comparing *beta* coefficients for the window identified as optimal by *climwin* Figs. `r all_traditional_comparisons`).

@@

The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_no_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
For PPT at LT, optimal seasonal windows pre-1970 and post-1970 were similar to those for the entire time period.
For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal seasonal window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig. `r SC_1970break`).
For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar seasonal windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig. `r SC_1970break`).
For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and seasonal windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
Further analysis into how climate sensitivities may be shifting would be valuable, but is beyond the scope of this analysis.

>>>>

traditional_comparison_generic_legend = "Shown are repsonses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis. Left panels show a month-by-month comparison of *beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable from species-level residual chronologies (traditional approach) and from individual-level analysis in *climwin* (approach presented here). Center panels compare the monthly *beta* coefficient estimates, with the dotted line indicating 1:1 correspondence. Finally, the right panels compare *beta* coefficients for the most influential window selected by *climwin*. Error bars indicate standard error of slope estimates. Note that 1:1 correspondence is not necessarily expected. See Appendix 5 for analysis methods and discussion of expected correspondence."

climwin_generic = "Rows correspond to the four different models run: (1) all RW data, (2) RW for the subset of cores for which $DBH could be reconstructed, (3) BAI, and (4)  AGB growth. The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1). Difference in $AIC$ relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column). The third and fourth columns give values of linear and quadratic terms for each seasonal window. The final column shows the correlation of individual-level residuals to the selected climate variable, with the function fit in *climwin* (not the final fit)."

@@

The Niobrara site sits at the prairie-forest ecotone of the North American Great Plains. 
There, forests occur in cool, moist canyons formed by spring-fed tributaries along the Niobrara River. 
*Betula papyrifera* occur in these canyons, in spots where there are groundwater springs that flow all year long.
Niobrara is near the southern range limit of this species, and populations at this site are small and highly disjunct.
Cores were taken from seven north-facing *Betula papyrifera* stands along a 27 km section of the Niobrara River, as detailed in @bumann_assessing_2019. 
The largest healthy trees within each stand were selected for coring.

@@

Prior to analysis, data were prepared and cleaned as described in the Methods section, resulting in an identical set of records for input into each analysis.
For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and seasonal windows can be compared without climwin [e.g., @meko_seascorr_2011], and *climwin* can also be used on residual chronologies. 
Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.
*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the seasonal window identified as most influential by *climwin*.

@@

Finally, we generated plots comparing month-by-month *beta* coefficients describing climate sensitivity, and also comparing *beta* coefficients for the window identified as most influential by *climwin* Figs. `r all_traditional_comparisons`).

@@

The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_no_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
For PPT at LT, the most influential seasonal windows pre-1970 and post-1970 were similar to those for the entire time period.
For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's most influential seasonal window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the most influential climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig. `r SC_1970break`).
For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar seasonal windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig. `r SC_1970break`).
For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and seasonal windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
Further analysis into how climate sensitivities may be shifting would be valuable, but is beyond the scope of this analysis.
